---
title: "grep - searching for words"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
```

Load the data
```{r}
osha <- read_csv("data/osha.csv")
```

Searching for keywords inside of columns:
such as "farming" or "fish" in industry_description.
```{r}
osha %>% filter(grepl("farming",industry_description)) %>% 
  count(industry_description)
```

Remember that R is case sensitive, and that's true here as well. But `grepl()` comes with an argument that ignores case: 
```{r}
osha %>% filter(grepl("farming",industry_description, ignore.case=TRUE)) %>% 
  count(industry_description)
```

Now try "fish":
```{r}
osha %>% filter(grepl("fish",industry_description, ignore.case=TRUE)) %>% 
  count(industry_description)
```
```{r}
?grep
```

The `grepl()` function is in a family of `grep` functions that allow you to search for strings (as above) or patters. In order to search for patterns you need to learn "regular expressions", which is a whole other class. But you can use `grep()` and `grepl()` to search for keywords in a text column. `grepl()` returns TRUE or FALSE depending on whether the text matches, and `grep()` returns an index of the values that match:

```{r}
a <- c("wheat farming", "soybean farming", "alfalfa farming", "cod fishing", "salmon fishing")
```

```{r}
grep("farm", a)
grepl("farm",a)
```

