---
title: "Salaries"
output: html_notebook
---

```{r}
library(tidyverse)
library(readxl)
```

Load data
```{r}
salaries <- read_excel("data/salaries.xlsx", sheet=1)
```

Run integrity checks on columns in your data

Use the `count()` function to look for misspellings and inconsistencies (and also to see which values are most common)
```{r}
salaries %>% count(department)
```

These two lines of code produce exactly the same results.
```{r}
salaries %>% count(job_title)
salaries %>% group_by(job_title) %>% summarise(n = n())
```

Which employee made the most amount of money (hint: total compensation)? 
```{r}
salaries %>% arrange(desc(total_comp))
```

What is the total compensation paid out by the city in 2021? 
```{r}
salaries %>% summarise(overtime = sum(overtime_oncall, na.rm=TRUE))
```

Which department paid out the most in overtime pay? 
```{r}
salaries %>% 
  group_by(department) %>% 
  summarise(overtime = sum(overtime_oncall, na.rm=TRUE)) %>% 
  arrange(desc(overtime))
```

Which job title received the most in overtime pay? 
```{r}
salaries %>% 
  group_by(job_title) %>% 
  summarise(OT = sum(overtime_oncall, na.rm=T)) %>% 
  arrange(desc(OT))
```

Explore which Senior Police Officers are getting the most overtime:
```{r}
salaries %>% filter(job_title=="Senior Police Officer") %>% 
  arrange(desc(overtime_oncall))
```

What's the most common first name? 
```{r}
salaries %>% count(first_name) %>% arrange(desc(n))
```

